---
title: "Poisson Regression and Maximum Likelihood Estimation"
subtitle: "A Case Study of Blueprinty's Software and Patent Awards"
author: "Jiachen Xu"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
---

```{r}
#| include: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(gridExtra)
library(glue)

# Read the data
blueprinty_data <- read.csv("blueprinty.csv")
```

## Introduction

Blueprinty is a small but ambitious company with a clear value proposition: they develop specialized software designed to help engineering firms navigate the patent application process more successfully. Their tools are built specifically to streamline blueprint preparation and submission to the U.S. Patent Office.

The marketing team at Blueprinty has an appealing story they'd like to tell: *firms using our software win more patent approvals*. But here's the challenge: proving this claim is surprisingly tricky.

In an ideal world, we'd have data showing each firm's patent success before and after adopting Blueprinty's software. But that data doesn't exist. What we do have is a snapshot of 1,500 mature engineering firms—established companies (not startups)—along with:

- How many patents each firm had approved over the last 5 years
- The firm's age since incorporation
- Geographic region
- Whether or not the firm uses Blueprinty's software

On the surface, the claim seems testable: just compare patent counts between customers and non-customers. But there's a fundamental problem lurking beneath the surface: **selection bias**. Blueprinty's customers didn't arrive randomly. Firms that chose to adopt the software might be fundamentally different from those that didn't. Perhaps larger, more established firms are more likely to invest in specialized software. Perhaps certain regions have stronger innovation cultures. Any raw difference in patent counts could reflect these unobserved differences in firm quality and capabilities rather than the actual impact of the software itself.

In this analysis, we'll use **Poisson regression** and **maximum likelihood estimation** to carefully estimate the effect of Blueprinty's software while accounting for measurable differences between customers and non-customers. We can't solve the fundamental selection bias problem, but we can make progress.


## Exploring the Data

Let's start by understanding what we're working with.

```{r}
#| echo: false

# Summary statistics
overview_table <- data.frame(
  Metric = c(
    "Dataset size",
    "Patent count range",
    "Blueprinty customers",
    "Mean firm age"
  ),
  Value = c(
    paste(nrow(blueprinty_data), "firms"),
    paste(min(blueprinty_data$patents), "to", max(blueprinty_data$patents)),
    paste0(sum(blueprinty_data$iscustomer), " firms (",
           round(100 * mean(blueprinty_data$iscustomer), 1), "%)"),
    paste(round(mean(blueprinty_data$age), 1), "years")
  )
)

knitr::kable(overview_table, row.names = FALSE)
```

### Patent Awards by Customer Status

First, let's compare the raw distribution of patent awards between firms using Blueprinty and those that don't.

```{r}
#| fig-width: 12
#| fig-height: 5
#| echo: false

# Calculate summary statistics by group
summary_stats <- blueprinty_data %>%
  group_by(iscustomer) %>%
  summarise(
    mean_patents = mean(patents),
    median_patents = median(patents),
    sd_patents = sd(patents),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    group = c("Non-customers", "Blueprinty Customers")[iscustomer + 1]
  )

n_non <- summary_stats$n[summary_stats$iscustomer == 0]
n_cust <- summary_stats$n[summary_stats$iscustomer == 1]
mean_non <- summary_stats$mean_patents[summary_stats$iscustomer == 0]
mean_cust <- summary_stats$mean_patents[summary_stats$iscustomer == 1]

# Create side-by-side histograms
p1 <- blueprinty_data %>%
  filter(iscustomer == 0) %>%
  ggplot(aes(x = patents)) +
  geom_histogram(bins = 15, fill = "#E8E8E8", color = "black", alpha = 0.8) +
  labs(
    title = glue::glue("Non-Customers (n = {n_non})"),
    x = "Patents Awarded (5-year period)",
    y = "Number of Firms"
  ) +
  coord_cartesian(xlim = c(-0.5, 15)) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    panel.grid.major.x = element_blank()
  )

p2 <- blueprinty_data %>%
  filter(iscustomer == 1) %>%
  ggplot(aes(x = patents)) +
  geom_histogram(bins = 15, fill = "#4A90E2", color = "black", alpha = 0.8) +
  labs(
    title = glue::glue("Blueprinty Customers (n = {n_cust})"),
    x = "Patents Awarded (5-year period)",
    y = "Number of Firms"
  ) +
  coord_cartesian(xlim = c(-0.5, 15)) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    panel.grid.major.x = element_blank()
  )

grid.arrange(p1, p2, nrow = 1)

# Print summary table
summary_stats %>%
  select(group, mean_patents, median_patents, sd_patents, n) %>%
  rename(
    Group = group,
    "Mean Patents" = mean_patents,
    "Median Patents" = median_patents,
    "Std Dev" = sd_patents,
    "N Firms" = n
  ) %>%
  knitr::kable(digits = 2, row.names = FALSE)
```

**What we observe:** Blueprinty customers have a notably higher mean number of patents (`r round(mean_cust, 2)` vs. `r round(mean_non, 2)`). On the surface, this supports the marketing claim—but we must be cautious. The customer group is smaller (`r n_cust` vs. `r n_non` firms), and the distribution shapes overlap substantially, suggesting confounding factors may be at play.

### Selection Bias: Firm Age and Geography

To understand potential selection bias, let's examine whether customers and non-customers differ systematically on observable characteristics.

```{r}
#| fig-width: 12
#| fig-height: 5
#| echo: false

# Compare firm ages
p_age_customer <- blueprinty_data %>%
  ggplot(aes(x = age, fill = factor(iscustomer))) +
  geom_histogram(bins = 20, alpha = 0.6, position = "dodge") +
  scale_fill_manual(
    values = c("0" = "#E8E8E8", "1" = "#4A90E2"),
    labels = c("0" = "Non-customers", "1" = "Customers")
  ) +
  labs(
    title = "Firm Age Distribution",
    x = "Years Since Incorporation",
    y = "Number of Firms",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = c(0.98, 0.97),
        legend.justification = c(1, 1))

# Age statistics
age_by_group <- blueprinty_data %>%
  group_by(iscustomer) %>%
  summarise(
    mean_age = mean(age),
    sd_age = sd(age),
    median_age = median(age),
    .groups = "drop"
  )

print(age_by_group)

mean_age_non <- age_by_group$mean_age[age_by_group$iscustomer == 0]
mean_age_cust <- age_by_group$mean_age[age_by_group$iscustomer == 1]

# Region distribution
region_crosstab <- table(blueprinty_data$region, blueprinty_data$iscustomer)
region_pct <- prop.table(region_crosstab, margin = 2) * 100

p_region <- blueprinty_data %>%
  group_by(region, iscustomer) %>%
  count(name = "count") %>%
  pivot_wider(names_from = iscustomer, values_from = count, values_fill = 0) %>%
  mutate(
    pct_non = 100 * `0` / sum(`0`),
    pct_customer = 100 * `1` / sum(`1`)
  ) %>%
  select(region, pct_non, pct_customer) %>%
  pivot_longer(
    cols = starts_with("pct"),
    names_to = "group",
    values_to = "pct"
  ) %>%
  mutate(group = c("pct_customer" = "Customers", "pct_non" = "Non-customers")[group]) %>%
  ggplot(aes(x = reorder(region, -pct), y = pct, fill = group)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(
    values = c("Non-customers" = "#E8E8E8", "Customers" = "#4A90E2")
  ) +
  labs(
    title = "Regional Distribution by Customer Status",
    x = "Region",
    y = "Percentage of Group",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = c(0.98, 0.97),
    legend.justification = c(1, 1)
  )

grid.arrange(p_age_customer, p_region, nrow = 1)
```

**Key findings on selection bias:**

- **Age:** Blueprinty customers are slightly older on average (`r round(mean_age_cust, 1)` years) compared to non-customers (`r round(mean_age_non, 1)` years). The gap is not dramatic, but it is still a reminder that customers and non-customers are not randomly assigned.

- **Geography:** The regional distribution differs sharply between groups. Customers are especially concentrated in the Northeast (where Blueprinty may have stronger sales coverage or product-market fit) and less represented across the other regions.

These differences matter because age and geography may independently influence patenting success. Simply comparing raw patent counts conflates the effect of the software with these underlying firm characteristics. That's where regression comes in.


## A Simple Poisson Model via Maximum Likelihood

### Why Poisson Regression?

Our outcome variable is a count of patents—a non-negative integer. The Normal distribution is a poor choice here because:

- It allows negative values (patents can't be negative)
- It assumes continuous variation (patents are discrete)
- It assumes constant variance (variance typically increases with the mean for counts)

The **Poisson distribution** is the natural starting point for modeling count data. It assumes that outcomes are non-negative integers and that the variance equals the mean—a property that often holds reasonably well for patent data.

### Deriving the Likelihood

The Poisson probability mass function is:

$$
f(Y|\lambda) = \frac{e^{-\lambda}\lambda^{Y}}{Y!}
$$

where $Y$ is the number of patents and $\lambda$ is the expected count.

For $n$ independent observations $(Y_1, Y_2, \ldots, Y_n)$, the **joint likelihood** is the product:

$$
L(\lambda) = \prod_{i=1}^{n} \frac{e^{-\lambda}\lambda^{Y_i}}{Y_i!}
$$

Taking the natural logarithm (which preserves the location of the maximum):

$$
\ell(\lambda) = \sum_{i=1}^{n} \log\left(\frac{e^{-\lambda}\lambda^{Y_i}}{Y_i!}\right)
$$

Expanding each term:

$$
\ell(\lambda) = \sum_{i=1}^{n} \left[ -\lambda + Y_i \log(\lambda) - \log(Y_i!) \right]
$$

Simplifying:

$$
\ell(\lambda) = -n\lambda + \log(\lambda)\sum_{i=1}^{n} Y_i - \sum_{i=1}^{n}\log(Y_i!)
$$

The last term doesn't depend on $\lambda$, so we can drop it for optimization purposes:

$$
\ell(\lambda) = -n\lambda + \sum_{i=1}^{n} Y_i \log(\lambda)
$$

### Coding and Visualizing the Log-Likelihood

```{r}
# Extract total patent count
Y_obs <- blueprinty_data$patents

# Define the Poisson log-likelihood function
poisson_loglikelihood <- function(lambda, Y) {
  n <- length(Y)
  sum_Y <- sum(Y)
  # Log-likelihood (up to a constant)
  loglik <- -n * lambda + sum_Y * log(lambda)
  return(loglik)
}

# Create a grid of lambda values and compute the log-likelihood
lambda_grid <- seq(0.1, 10, by = 0.05)
loglik_values <- sapply(lambda_grid, function(lam) poisson_loglikelihood(lam, Y_obs))

# Find the maximum
max_idx <- which.max(loglik_values)
mle_lambda_analytical <- lambda_grid[max_idx]
max_loglik <- loglik_values[max_idx]

# Plot the log-likelihood
ggplot(data.frame(lambda = lambda_grid, loglik = loglik_values), aes(x = lambda, y = loglik)) +
  geom_line(linewidth = 1, color = "#2E86DE") +
  geom_point(aes(x = mle_lambda_analytical, y = max_loglik), 
             size = 4, color = "#A23B72") +
  labs(
    title = "Poisson Log-Likelihood Function",
    subtitle = "The peak shows where the data is most consistent with a given λ",
    x = "λ (Expected number of patents)",
    y = "Log-Likelihood"
  ) +
  annotate(
    "text",
    x = mle_lambda_analytical + 0.5,
    y = max_loglik - 50,
    label = glue::glue("MLE: λ = {round(mle_lambda_analytical, 3)}"),
    fontface = "bold",
    color = "#A23B72"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11, color = "gray40")
  )
```

**Interpreting the curve:** The log-likelihood curve peaks at the value of $\lambda$ that makes our observed data most likely. We can "read" the MLE visually off this curve—it's roughly `r round(mle_lambda_analytical, 2)` in this case.

### Analytical Solution

Taking the derivative of the log-likelihood and setting it equal to zero:

$$
\frac{d\ell}{d\lambda} = -n + \frac{1}{\lambda}\sum_{i=1}^{n} Y_i = 0
$$

Solving for $\lambda$:

$$
\hat{\lambda}_{\text{MLE}} = \frac{1}{n}\sum_{i=1}^{n} Y_i = \bar{Y}
$$

This is elegant and intuitive: **the MLE of $\lambda$ is simply the sample mean**. For our data, that's:

```{r}
sample_mean <- mean(Y_obs)
cat(sprintf("Sample mean (analytical MLE): %.4f\n", sample_mean))
cat(sprintf("Log-likelihood at sample mean: %.2f\n", poisson_loglikelihood(sample_mean, Y_obs)))
```

### Numerical Optimization with optim()

Let's verify this with numerical optimization:

```{r}
# Use optim() to find the MLE
# Note: optim minimizes by default, so we negate the log-likelihood
mle_result <- optim(
  par = 3,  # Initial guess
  fn = function(lam) -poisson_loglikelihood(lam, Y_obs),  # Negative because optim minimizes
  method = "Brent",
  lower = 0.01,
  upper = 20
)

mle_lambda_numerical <- mle_result$par
cat(sprintf("Numerical MLE: %.4f\n", mle_lambda_numerical))
cat(sprintf("Analytical MLE (sample mean): %.4f\n", sample_mean))
cat(sprintf("Difference: %.6f\n", abs(mle_lambda_numerical - sample_mean)))
```

**Comparison:** The numerical optimization recovers essentially the same answer as the analytical solution (the tiny difference is just numerical rounding). This isn't a coincidence—it's because we have a convex optimization problem with a unique global maximum.


## Poisson Regression

### Extension to Multiple Covariates

The simple model above assumes all firms have the same patent rate $\lambda$. But we know firms differ by age, region, and customer status. In **Poisson regression**, we let each firm's rate parameter depend on its own characteristics:

$$
Y_i \sim \text{Poisson}(\lambda_i) \quad \text{where} \quad \lambda_i = \exp(X_i'\beta)
$$

Here, $X_i$ is a vector of firm characteristics (including an intercept), and $\beta$ is a vector of coefficients to estimate.

We use the exponential link function $\exp(\cdot)$ because:

- $\lambda_i$ must be positive (you can't have a negative expected count of patents)
- The linear predictor $X_i'\beta$ can be any real number
- The exponential transformation maps $(-\infty, \infty) \to (0, \infty)$, keeping $\lambda_i$ in the valid range

This is a **multiplicative model**: when we increase a covariate by 1 unit, the expected count gets multiplied by $e^{\beta_j}$, rather than increased by $\beta_j$.

### Setting up the Regression Model

```{r}
#| include: false

# Construct the design matrix X
# Columns: intercept, age, age^2, region dummies (dropping one), customer indicator

# Create region dummies (drop "Midwest" as reference)
X <- model.matrix(
  ~ age + I(age^2) + factor(region) + factor(iscustomer),
  data = blueprinty_data
)

# Remove the "factor(iscustomer)0" column (keep only the 1 indicator)
X <- X[, colnames(X) != "factor(iscustomer)0"]
colnames(X)

cat("Design matrix dimensions:", nrow(X), "x", ncol(X), "\n")
head(X, 3)
```

**Model specification:**

- **Intercept:** Baseline patent rate for a non-customer firm of age 0 in the Midwest
- **Age:** Linear effect of firm age on patent rate
- **Age²:** Allows the relationship between age and patenting to be non-linear (firms might innovate more heavily at middle age)
- **Region dummies:** Captures regional variation in patenting culture (Northeast, Northwest, South, Southwest; Midwest is reference)
- **Customer indicator:** The main effect of interest—how does being a Blueprinty customer affect patent rate?

### Regression Log-Likelihood

For the regression model, each observation has its own Poisson rate:

$$
\lambda_i = \exp(X_i'\beta)
$$

Substituting this into the Poisson probability mass function gives:

$$
L(\beta) = \prod_{i=1}^{n} \frac{e^{-\exp(X_i'\beta)}\left[\exp(X_i'\beta)\right]^{Y_i}}{Y_i!}
$$

Taking logs:

$$
\ell(\beta) = \sum_{i=1}^{n} \left[-\exp(X_i'\beta) + Y_i X_i'\beta - \log(Y_i!)\right]
$$

The final term, $\log(Y_i!)$, does not depend on $\beta$, so it can be dropped for numerical optimization. That leaves the objective we code below:

$$
\ell(\beta) \propto \sum_{i=1}^{n} \left[-\exp(X_i'\beta) + Y_i X_i'\beta\right]
$$

### Maximum Likelihood Estimation with optim()

```{r}
# Define the Poisson regression log-likelihood
poisson_regression_loglikelihood <- function(beta, Y, X) {
  # Linear predictor: X * beta
  eta <- drop(X %*% beta)
  # Inverse link: exp(eta)
  lambda <- exp(eta)
  # Log-likelihood
  loglik <- sum(-lambda + Y * eta)
  return(loglik)
}

# Gradient of the negative log-likelihood. Supplying this helps optim()
# converge reliably with age and age-squared in the same design matrix.
poisson_regression_gradient <- function(beta, Y, X) {
  eta <- drop(X %*% beta)
  lambda <- exp(eta)
  -drop(t(X) %*% (Y - lambda))
}

# Fit the model with optim()
# Start with zeros as initial guess
initial_beta <- rep(0, ncol(X))

# Use BFGS with the analytic gradient
mle_fit <- optim(
  par = initial_beta,
  fn = function(b) -poisson_regression_loglikelihood(b, Y_obs, X),
  gr = function(b) poisson_regression_gradient(b, Y_obs, X),
  method = "BFGS",
  hessian = FALSE,
  control = list(maxit = 10000, reltol = 1e-12)
)

beta_mle <- mle_fit$par

# Compute standard errors from the analytic information matrix.
lambda_mle <- as.numeric(exp(X %*% beta_mle))
hessian_matrix <- t(X) %*% (X * lambda_mle)
information_matrix <- solve(hessian_matrix)
se_beta <- sqrt(diag(information_matrix))

# Create results table
results_df <- data.frame(
  Variable = colnames(X),
  Coefficient = beta_mle,
  StdErr = se_beta,
  z_stat = beta_mle / se_beta,
  p_value = 2 * (1 - pnorm(abs(beta_mle / se_beta)))
)

# Display results
cat("MLE Results for Poisson Regression\n")
cat("Log-likelihood:", -mle_fit$value, "\n\n")

results_table <- results_df %>%
  mutate(
    Coefficient = round(Coefficient, 4),
    StdErr = round(StdErr, 4),
    z_stat = round(z_stat, 3),
    p_value = round(p_value, 4)
  ) %>%
  rename(
    "Coef." = Coefficient,
    "Std. Error" = StdErr,
    "z-statistic" = z_stat,
    "p-value" = p_value
  ) %>%
  knitr::kable(digits = 4, row.names = FALSE)

results_table
```

### Comparison with glm()

Let's verify our results by fitting the same model with R's built-in `glm()` function:

```{r}
# Fit with glm
glm_model <- glm(
  patents ~ age + I(age^2) + factor(region) + factor(iscustomer),
  family = poisson(link = "log"),
  data = blueprinty_data
)

# Extract coefficients and standard errors
glm_coef <- coef(glm_model)
glm_se <- summary(glm_model)$coefficients[, "Std. Error"]

# Compare
comparison_df <- data.frame(
  Variable = names(glm_coef),
  MLE_Coef = beta_mle,
  GLM_Coef = as.numeric(glm_coef),
  MLE_SE = se_beta,
  GLM_SE = glm_se
) %>%
  mutate(
    Coef_Diff = abs(MLE_Coef - GLM_Coef),
    SE_Diff = abs(MLE_SE - GLM_SE)
  )

comparison_tbl <- comparison_df %>%
  select(Variable, MLE_Coef, GLM_Coef, Coef_Diff) %>%
  mutate(across(c(MLE_Coef, GLM_Coef, Coef_Diff), ~round(., 4))) %>%
  rename(
    "MLE Coefficient" = MLE_Coef,
    "GLM Coefficient" = GLM_Coef,
    "Difference" = Coef_Diff
  ) %>%
  knitr::kable(digits = 4, row.names = FALSE)

comparison_tbl
```

**Excellent agreement:** The MLE estimates from our hand-coded optimization match the GLM results almost exactly. That is a useful validation check: the likelihood we coded and the likelihood used by `glm()` are solving the same statistical problem.

### Interpreting the Results

```{r}
#| echo: false

# Create an interpretable results summary
interpret_df <- results_df %>%
  mutate(
    exp_coef = exp(Coefficient),
    pct_change = (exp_coef - 1) * 100,
    sig = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(Variable, Coefficient, exp_coef, pct_change, sig, p_value)

interpret_df %>%
  mutate(
    Coefficient = round(Coefficient, 4),
    exp_coef = round(exp_coef, 4),
    pct_change = round(pct_change, 2),
    p_value = round(p_value, 4)
  ) %>%
  rename(
    "β" = Coefficient,
    "exp(β)" = exp_coef,
    "% Change" = pct_change,
    "Signif." = sig,
    "p-value" = p_value
  ) %>%
  knitr::kable(digits = 4, row.names = FALSE)
```

**Key interpretations:**

- **Age:** Positive coefficient suggests more established firms get more patents. This makes sense—older, larger firms have more resources for R&D.

- **Age²:** Negative coefficient indicates diminishing returns to age. Very old firms may plateau in their innovation.

- **Region effects:** Once age and customer status are included, the regional coefficients are comparatively small and not very precisely estimated. Geography still matters for selection into Blueprinty usage, but it is not the main driver of patent counts in this specification.

- **Blueprinty Customer:** The coefficient on `factor(iscustomer)1` is our main estimate of interest. A positive coefficient would suggest customers get more patents even after controlling for age and region.

### Model Diagnostics and Goodness of Fit

How well does our model fit the data? Let's examine predicted vs. actual patent counts.

```{r}
#| fig-width: 12
#| fig-height: 5
#| echo: false

# Get fitted values from our model
fitted_lambda <- exp(X %*% beta_mle)

# Create diagnostic data frame
diagnostics <- data.frame(
  actual = Y_obs,
  predicted = as.numeric(fitted_lambda),
  residuals = Y_obs - as.numeric(fitted_lambda),
  pearson_residuals = (Y_obs - as.numeric(fitted_lambda)) / sqrt(as.numeric(fitted_lambda))
)

# Plot 1: Actual vs Predicted
p_fit1 <- ggplot(diagnostics, aes(x = predicted, y = actual)) +
  geom_point(alpha = 0.5, color = "#4A90E2") +
  geom_abline(intercept = 0, slope = 1, color = "#E74C3C", linewidth = 1, linetype = "dashed") +
  labs(
    title = "Actual vs Predicted Patent Counts",
    x = "Predicted Count (λ)",
    y = "Actual Count"
  ) +
  theme_minimal()

# Plot 2: Pearson Residuals vs Predicted
p_fit2 <- ggplot(diagnostics, aes(x = predicted, y = pearson_residuals)) +
  geom_point(alpha = 0.5, color = "#27AE60") +
  geom_hline(yintercept = 0, color = "#E74C3C", linewidth = 1, linetype = "dashed") +
  geom_hline(yintercept = c(-2, 2), color = "gray", linetype = "dotted") +
  labs(
    title = "Residual Plot: Checking Model Assumptions",
    x = "Predicted Count (λ)",
    y = "Pearson Residuals"
  ) +
  theme_minimal()

grid.arrange(p_fit1, p_fit2, nrow = 1)

# Goodness of fit statistics
cat("Model Fit Statistics:\n")
cat("Mean Absolute Error (MAE):", round(mean(abs(diagnostics$residuals)), 3), "\n")
cat("Root Mean Squared Error (RMSE):", round(sqrt(mean(diagnostics$residuals^2)), 3), "\n")
cat("Residual deviance:", round(glm_model$deviance, 2), "\n")
cat("AIC:", round(AIC(glm_model), 2), "\n")
```

The residual plot shows relatively balanced scatter around zero, suggesting our Poisson assumption is reasonable. The actual vs. predicted plot confirms the model captures the general trend in patent counts.

### Coefficient Visualization: Forest Plot

Visual representation of all coefficients with confidence intervals provides a more intuitive understanding:

```{r}
#| fig-width: 10
#| fig-height: 6
#| echo: false

# Compute 95% confidence intervals
ci_lower <- beta_mle - 1.96 * se_beta
ci_upper <- beta_mle + 1.96 * se_beta

# Create forest plot data
forest_data <- data.frame(
  Variable = colnames(X),
  Coefficient = beta_mle,
  CI_Lower = ci_lower,
  CI_Upper = ci_upper,
  Significant = ifelse(ci_lower * ci_upper > 0, "Yes", "No")
) %>%
  arrange(Coefficient)

# Create forest plot
ggplot(forest_data, aes(y = reorder(Variable, Coefficient), x = Coefficient)) +
  geom_vline(xintercept = 0, color = "gray", linetype = "dashed", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper, color = Significant), 
                 height = 0.2, linewidth = 1) +
  geom_point(aes(color = Significant), size = 3) +
  scale_color_manual(
    values = c("Yes" = "#E74C3C", "No" = "#95A5A6"),
    name = "Statistically\nSignificant"
  ) +
  labs(
    title = "Forest Plot: Poisson Regression Coefficients with 95% CI",
    x = "Coefficient Value",
    y = "Variable"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

The forest plot reveals which variables significantly influence patent counts. The wide confidence intervals around age-squared reflect the relatively small effect and higher uncertainty in that parameter.


## The Effect of Blueprinty's Software

Now for the main question: **How much do Blueprinty customers get in patent awards compared to similar non-customers?**

Because our model is multiplicative (not additive), we can't simply interpret the regression coefficient. Instead, we use a **counterfactual analysis**: we'll imagine two scenarios for each firm and compare predicted outcomes.

### Counterfactual Predictions

```{r}
#| echo: false

# Create two counterfactual datasets:
# X0: Every firm as a non-customer (iscustomer = 0)
# X1: Every firm as a customer (iscustomer = 1)

X0 <- X
X0[, "factor(iscustomer)1"] <- 0

X1 <- X
X1[, "factor(iscustomer)1"] <- 1

# Predict expected patents under each scenario
pred_lambda_0 <- exp(X0 %*% beta_mle)
pred_lambda_1 <- exp(X1 %*% beta_mle)

# Individual treatment effects
individual_effects <- pred_lambda_1 - pred_lambda_0

# Average treatment effect (ATE)
ate <- mean(individual_effects)

cat(sprintf("Average Treatment Effect: %.4f additional patents\n", ate))
cat(sprintf("As a percentage increase: %.2f%%\n", (ate / mean(pred_lambda_0)) * 100))
```

### Visualizing the Effect

```{r}
#| fig-width: 10
#| fig-height: 6
#| echo: false

# Create a data frame for visualization
effect_data <- data.frame(
  firm_id = 1:nrow(blueprinty_data),
  prediction_non = as.numeric(pred_lambda_0),
  prediction_customer = as.numeric(pred_lambda_1),
  effect = individual_effects,
  age = blueprinty_data$age,
  region = blueprinty_data$region
)

# Plot 1: Distribution of treatment effects
p1 <- ggplot(effect_data, aes(x = effect)) +
  geom_histogram(bins = 30, fill = "#4A90E2", alpha = 0.7, color = "black") +
  geom_vline(aes(xintercept = ate), color = "#E74C3C", linewidth = 1, linetype = "dashed") +
  labs(
    title = "Distribution of Individual Treatment Effects",
    x = "Predicted Additional Patents (Customer vs Non-Customer)",
    y = "Number of Firms"
  ) +
  annotate(
    "text",
    x = ate + 0.1,
    y = Inf,
    label = glue::glue("Average: {round(ate, 3)}"),
    vjust = 1.5,
    hjust = 0,
    color = "#E74C3C",
    fontface = "bold"
  ) +
  theme_minimal()

# Plot 2: Effect by firm age
p2 <- effect_data %>%
  mutate(age_bin = cut(age, breaks = 5)) %>%
  group_by(age_bin) %>%
  summarise(mean_effect = mean(effect), .groups = "drop") %>%
  ggplot(aes(x = age_bin, y = mean_effect)) +
  geom_col(fill = "#27AE60", alpha = 0.8, color = "black") +
  labs(
    title = "Treatment Effect by Firm Age Group",
    x = "Firm Age (years)",
    y = "Avg. Additional Patents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1, p2, nrow = 1)
```

### Summary of Findings

```{r}
#| echo: false

summary_table <- data.frame(
  Statistic = c(
    "Average Treatment Effect",
    "Percentage increase",
    "Median Effect",
    "Min Individual Effect",
    "Max Individual Effect",
    "Std Dev of Effect"
  ),
  Value = c(
    round(ate, 4),
    round((ate / mean(pred_lambda_0)) * 100, 2),
    round(median(individual_effects), 4),
    round(min(individual_effects), 4),
    round(max(individual_effects), 4),
    round(sd(individual_effects), 4)
  )
) %>%
  rename(Estimate = Statistic) %>%
  knitr::kable(digits = 4, row.names = FALSE)

summary_table
```

### Confidence Intervals for the Treatment Effect

To quantify uncertainty around the ATE, we can use the bootstrap method or a parametric approach. Here's a bootstrap-based 95% confidence interval:

```{r}
#| echo: false

# Bootstrap confidence interval for ATE
set.seed(42)
n_bootstrap <- 1000
bootstrap_ates <- rep(NA, n_bootstrap)

for (i in 1:n_bootstrap) {
  # Resample with replacement
  idx_boot <- sample(1:nrow(blueprinty_data), replace = TRUE)
  Y_boot <- Y_obs[idx_boot]
  X_boot <- X[idx_boot, ]
  
  # Fit model on bootstrap sample
  tryCatch({
    mle_boot <- optim(
      par = beta_mle,
      fn = function(b) -poisson_regression_loglikelihood(b, Y_boot, X_boot),
      gr = function(b) poisson_regression_gradient(b, Y_boot, X_boot),
      method = "BFGS",
      control = list(maxit = 1000, reltol = 1e-10)
    )
    
    # Compute ATE on bootstrap sample
    X0_boot <- X_boot
    X0_boot[, "factor(iscustomer)1"] <- 0
    
    X1_boot <- X_boot
    X1_boot[, "factor(iscustomer)1"] <- 1
    
    bootstrap_ates[i] <- mean(exp(X1_boot %*% mle_boot$par)) - 
                         mean(exp(X0_boot %*% mle_boot$par))
  }, error = function(e) {
    bootstrap_ates[i] <<- NA
  })
}

# Remove NAs and compute CI
bootstrap_ates_clean <- bootstrap_ates[!is.na(bootstrap_ates)]
ci_lower_ate <- quantile(bootstrap_ates_clean, 0.025)
ci_upper_ate <- quantile(bootstrap_ates_clean, 0.975)

cat("Average Treatment Effect Point Estimate:", round(ate, 4), "\n")
cat("95% Bootstrap Confidence Interval: [", 
    round(ci_lower_ate, 4), ", ", 
    round(ci_upper_ate, 4), "]\n")

# Visualize bootstrap distribution
ggplot(data.frame(ATE = bootstrap_ates_clean), aes(x = ATE)) +
  geom_histogram(bins = 40, fill = "#4A90E2", alpha = 0.7, color = "black") +
  geom_vline(aes(xintercept = ate), color = "#E74C3C", linewidth = 1.5, linetype = "solid") +
  geom_vline(aes(xintercept = ci_lower_ate), color = "#F39C12", linewidth = 1, linetype = "dashed") +
  geom_vline(aes(xintercept = ci_upper_ate), color = "#F39C12", linewidth = 1, linetype = "dashed") +
  labs(
    title = "Bootstrap Distribution of Average Treatment Effect",
    x = "Estimated ATE (additional patents)",
    y = "Frequency"
  ) +
  annotate(
    "text",
    x = ate,
    y = Inf,
    label = glue::glue("Point Est: {round(ate, 3)}\n95% CI: [{round(ci_lower_ate, 3)}, {round(ci_upper_ate, 3)}]"),
    vjust = 1.5,
    hjust = 0,
    size = 3.5,
    color = "#E74C3C",
    fontface = "bold"
  ) +
  theme_minimal()
```

The confidence interval reflects genuine uncertainty about the true treatment effect. The fact that it doesn't include zero suggests the effect is meaningful, but we must still account for potential unmeasured confounding.


## Robustness Checks: Model Sensitivity

Do our conclusions change if we modify the model specification? Testing robustness is crucial for credibility.

```{r}
#| echo: false

# Model 1: Original (full model)
model_full <- glm(
  patents ~ age + I(age^2) + factor(region) + factor(iscustomer),
  family = poisson(link = "log"),
  data = blueprinty_data
)

# Model 2: No age squared (linear age effect)
model_no_age2 <- glm(
  patents ~ age + factor(region) + factor(iscustomer),
  family = poisson(link = "log"),
  data = blueprinty_data
)

# Model 3: No region controls
model_no_region <- glm(
  patents ~ age + I(age^2) + factor(iscustomer),
  family = poisson(link = "log"),
  data = blueprinty_data
)

# Model 4: Simple model (only customer effect)
model_simple <- glm(
  patents ~ factor(iscustomer),
  family = poisson(link = "log"),
  data = blueprinty_data
)

# Extract customer coefficients and their exponentiated forms
robustness_check <- data.frame(
  Model = c(
    "Full Model\n(age, age², region, customer)",
    "No age²\n(age, region, customer)",
    "No region\n(age, age², customer)",
    "Simple Model\n(customer only)"
  ),
  Customer_Coef = c(
    coef(model_full)["factor(iscustomer)1"],
    coef(model_no_age2)["factor(iscustomer)1"],
    coef(model_no_region)["factor(iscustomer)1"],
    coef(model_simple)["factor(iscustomer)1"]
  ),
  Effect_Multiplier = c(
    exp(coef(model_full)["factor(iscustomer)1"]),
    exp(coef(model_no_age2)["factor(iscustomer)1"]),
    exp(coef(model_no_region)["factor(iscustomer)1"]),
    exp(coef(model_simple)["factor(iscustomer)1"])
  ),
  Pct_Increase = c(
    (exp(coef(model_full)["factor(iscustomer)1"]) - 1) * 100,
    (exp(coef(model_no_age2)["factor(iscustomer)1"]) - 1) * 100,
    (exp(coef(model_no_region)["factor(iscustomer)1"]) - 1) * 100,
    (exp(coef(model_simple)["factor(iscustomer)1"]) - 1) * 100
  )
)

robustness_tbl <- robustness_check %>%
  mutate(
    Customer_Coef = round(Customer_Coef, 4),
    Effect_Multiplier = round(Effect_Multiplier, 4),
    Pct_Increase = round(Pct_Increase, 2)
  ) %>%
  rename(
    "Model Specification" = Model,
    "Customer β" = Customer_Coef,
    "exp(β)" = Effect_Multiplier,
    "% Increase" = Pct_Increase
  ) %>%
  knitr::kable(digits = 4, row.names = FALSE)

robustness_tbl

cat("\nInterpretation of Robustness Checks:\n")
cat("- All model specifications show a positive customer effect\n")
cat("- The full model estimates a larger effect than the simple comparison, mainly after modeling the age pattern flexibly\n")
cat("- Removing region controls barely changes the customer coefficient in this specification\n")
cat("- The full model is still preferred because it accounts for the observable differences we can measure\n")
```

### Key Findings from Sensitivity Analysis

**The robustness checks confirm three important patterns:**

1. **Direction is robust:** Every specification shows a positive effect of being a Blueprinty customer. This is reassuring—the conclusion doesn't hinge on a specific modeling choice.

2. **Magnitude depends on controls:** The simple model (without controls) implies about a `r round(robustness_check$Pct_Increase[robustness_check$Model == "Simple Model\n(customer only)"], 1)`% customer lift, while the full model implies about a `r round(robustness_check$Pct_Increase[robustness_check$Model == "Full Model\n(age, age², region, customer)"], 1)`% lift. The adjustment does not erase the customer effect; if anything, modeling the nonlinear age pattern makes the estimated association stronger.

3. **Geography matters for adoption, but less for this coefficient:** Customers are disproportionately in the Northeast, but removing region controls barely changes the customer estimate. In this dataset, age adjustment matters more for the customer coefficient than regional adjustment.

This pattern is **exactly what we'd expect** from a well-specified model that correctly handles observable confounding.

## Conclusion: What Can We Claim?

Our analysis provides evidence **consistent with** Blueprinty's marketing claim. After controlling for firm age and geographic region, the model suggests that being a Blueprinty customer is associated with approximately **`r round(ate, 2)` additional patents** in a 5-year period.

### Important Caveats

However, we must be very careful about interpreting causality here:

1. **This is an observational study, not a randomized experiment.** Firms selected into being Blueprinty customers; they weren't randomly assigned to use the software. Even though we've controlled for age and region, there may be unmeasured differences between customers and non-customers that drive the estimated effect.

2. **Selection bias is likely.** Blueprinty's most sophisticated customers—those who invest in patent optimization software—might also invest more heavily in R&D generally, have better engineering talent, or have stronger product-market fit. These unobserved advantages could drive higher patent approval rates independently of the software.

3. **The estimated effect could be an overestimate.** If customers are systematically higher-quality firms in unobserved ways, the model attributes their patent success to the software when it may reflect inherent firm strength.

4. **Reverse causality?** Perhaps firms that are already successful at patenting are more likely to invest in Blueprinty's software, rather than the software causing the success.

### What We've Learned

Despite these limitations, this analysis has value:

- **It corrects for observable confounders** (age and region), accounting for systematic differences between customer and non-customer populations.
- **It demonstrates a methodology** (MLE and Poisson regression) that can be applied in many marketing contexts where selection bias is a concern.
- **It provides a baseline estimate** that future randomized trials or natural experiments could test.
- **It quantifies uncertainty** through standard errors, showing us the precision of our estimates.

For Blueprinty to make a truly compelling case about their software's causal impact, they would ideally conduct a randomized trial or find a natural experiment (like a sudden change in pricing) that exogenously shifted adoption.

---

*This analysis was conducted using R's GLM framework with Poisson regression, combined with manual maximum likelihood estimation for validation and pedagogical clarity.*
